---
title: "TidyTuesday Week 42: The Datasaurus Dozen"
date: 2020-10-13
author: joleenbelinda
output: html_notebook
---

The data this week comes from Alberto Cairo in his original [blog post](http://www.thefunctionalart.com/2016/08/download-datasaurus-never-trust-summary.html) published in 2016. Later, an R package was created by Steph Locke + Lucy McGowan:[`datasauRus`](https://cran.r-project.org/web/packages/datasauRus/vignettes/Datasaurus.html) containing the Datasaurus Dozen dataset.

## Goals for this Analysis

The golden purpose of the Datasaurus dataset is to teach the importance of visualising data rather than only reporting on summary statistics. 

I am using this opportunity to learn how to create an animation that cycles through each plot using `gganimate`.


## Data Preparation

```{r setup}

library(tidyverse)
library(ggthemes)
library(gganimate)

```

```{r importData}
tuesdata <- tidytuesdayR::tt_load(2020, week = 42)
datasaurus <- tuesdata$datasaurus
```

```{r viewData}
datasaurus
```
## First look: summaries

First, we are going to have a look at the summary statistics for the data

```{r summary}

datasaurus %>%
  group_by(dataset) %>%
  summarise(mean_x    = mean(x),
            mean_y    = mean(y),
            sd_x = sd(x),
            sd_y = sd(y),
            corr  = cor(x, y)
    )

```

## Second look: visualisation

Second, we plot each dataset as a scatterplot.

```{r scatterplots, fig.height=7, fig.width=5}
ggplot(datasaurus, aes(x = x, y = y)) +
  facet_wrap(~ dataset, ncol = 3) +
  geom_point() +
  labs(title="Scatterplot of 13 Datasets with Similar Summary Statistics",caption="TidyTuesday: Datasaurus Dozen") +
  theme_pander()
```
## Animate the plots

Creating an animation with `gganimate` is incredibly easy - you create your `ggplot` scatterplot as normal and then add the option for which states you want the animation to transition between.

```{r animation}
dino_animation <- ggplot(datasaurus, aes(x = x, y = y)) +
  geom_point() +
  theme_pander() +
  transition_states(dataset) +
  ease_aes('linear') +
  labs(title = "Datasaurus Dozen Scatterplots", subtitle = "Dataset: {closest_state}", caption="TidyTuesday: Datasaurus Dozen")

# Save the animation
anim_save(here::here("2020", "week 42", "output", "dinosaurus_animation.gif"))
```

