---
title: "TidyTuesday Week 41: NCAA Women's Basketball Tournament"
date: 2020-10-06
author: joleenbelinda
output: html_notebook
---

The TidyTuesday data this week comes from [FiveThirtyEight]{https://fivethirtyeight.com/features/louisiana-tech-was-the-uconn-of-the-80s/}. 

## Useful info from the FiveThirtyEight article

We can use NCAA Tournament seeds as a proxy for team strength over time.

This doesn’t explicitly factor in how a team performs in the tournament itself, but it does measure the general quality of a team’s entire season — plus, higher seeds tend to win more in the tournament anyway. 

To measure this, we awarded “seed points” in proportion to a given seed number’s expected wins in the tournament, calibrated to a 100-point scale where the No. 1 seed gets 100 points, No. 2 gets 70 points, and so forth.

## Goals for this analysis

Considering each row to be each school's results per conference per year. 

What are the average seed points (at the end of the tournament) if a school plays their first game at home compared to if they played their first game away? How has this changed over the years?

Is there a difference in the win rate if the first game is played at the home court compared to if it is played away? How has this changed over the years?

## Data Preparation

```{r setup}

library(tidyverse)
library(scales)
library(ggthemes)

```

```{r importData}

tuesdata <- tidytuesdayR::tt_load(2020, week = 41)

tournament <- tuesdata$tournament

```

Calculate seed points according to the table given by FiveThirtyEight.

```{r calculatePoints}
tournament_points <- tournament %>%
  mutate(seed = case_when(is.na(seed) == TRUE ~ 0,
                          TRUE ~ as.numeric(.$seed))) %>%
  mutate(points = case_when(seed == 1 ~ 100,
                            seed == 2 ~ 72.7,
                            seed == 3 ~ 54.5,
                            seed == 4 ~ 48.5,
                            seed == 5 ~ 33.3,
                            seed == 6 ~ 33.3,
                            seed == 7 ~ 27.3,
                            seed == 8 ~ 21.2,
                            seed == 9 ~ 18.2,
                            seed == 10 ~ 18.2,
                            seed == 11 ~ 18.2,
                            seed == 12 ~ 15.2,
                            seed == 13 ~ 9.09,
                            seed == 14 ~ 6.06,
                            seed == 15 ~ 3.03,
                            seed == 16 ~ 0,
                            TRUE ~ 0)) %>%
  mutate(seed_points = seed*points)

# View the data
tournament_points

```

Clean up some dirty data in the `x1st_game_at_home` variable

```{r dataCleaning}
tournament_points <- tournament_points %>%
  mutate(x1st_game_at_home = case_when(x1st_game_at_home == "Y^" ~ "Home",
                                       x1st_game_at_home == "Y" ~ "Home",
                                       x1st_game_at_home == "N" ~ "Away"))
```

## Data Visualisation

Calculate average points obtained per year, split by home/away on first game. 

The scatter plot shows that prior to 1995, away games awarded higher average seed points. However, after 1995 it seemed to become a more determining factor to the number of tournaments won if the first game was played on the home court.

```{r avgPointsByYear}
avgPointsByYear <- tournament_points %>%
  group_by(year, x1st_game_at_home) %>%
  summarise(meanSeedPoints = mean(seed_points))
```


```{r plot avgPointsByYear}
# Scatter plot
ggplot(avgPointsByYear, aes(x = year, y = meanSeedPoints)) +
  geom_point() +
  facet_wrap(~ x1st_game_at_home) +
  labs(title="Average seed points per year split by home/away on first game",y="Average Seed Points", x="Year",caption="TidyTuesday: NCAA Women's Basketball Tournament") +
  theme_pander() +
  theme(strip.text.x = element_text(size = 14),
        axis.title.y = element_text(margin = margin(20, 15, 20, 15)),
        panel.background = element_rect(fill = NA, color = "black"))
```

Calulcate the win rate per year split by home/away on first game.

The scatter plot shows that playing the first game at home has a big impact on the overall win rate in tournaments. There is an improvement in the win rate around 2005-2010 for first games played away.

```{r TourneyWinRate}
TourneyWinRate <- tournament_points %>%
  group_by(year, x1st_game_at_home) %>%
  summarise(winRate = sum(tourney_w)/(sum(tourney_w)+sum(tourney_l)))
```


```{r plot TourneyWinRate}
# Barplot
ggplot(TourneyWinRate, aes(x = year, y = winRate)) +
  geom_point() +
  facet_wrap(~ x1st_game_at_home) +
  scale_y_continuous(labels = label_percent(accuracy = 1), n.breaks = 8) +
  labs(title="Win rate per year split by home/away on first game",y="Win Rate", x="Year",caption="TidyTuesday: NCAA Women's Basketball Tournament") +
  theme_pander() +
  theme(strip.text.x = element_text(size = 14),
        axis.title.y = element_text(margin = margin(20, 15, 20, 15)),
        panel.background = element_rect(fill = NA, color = "black"))
```


